### 2010-07-28 Thesis/Chapter2.tex
- Update 28 July - Read chapter 2 and corrected spellings, transitions etc. + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@136 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Background in continuum mechanics for soft-tissue modelling}
\label{chap:softtissue}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organs' deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore a substantial comprehension of the mechanics involved and a knowledge of physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly expose their experimental validation.
\end{shortAbstract}

\section{Introduction}
In our everyday life, matter appears smooth and continuous: from the wood used to build your desk to the water you drink. But this is just illusion. The concept that matter is composed of discrete units has been around for millennia. In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century \citep{Lautrup05}. However, certain physical phenomena can be predicted with theories that pay no attention to the molecular structure of materials. Consider for instance the deformation of the horizontal board of a bookshelf under the weight of the books. The bending of the shelf can be modelled without considering its molecular composition. The branch of physics in which materials are treated as continuous is known as \emph{continuum mechanics}. Continuum mechanics studies the response of materials to different loading conditions. In this theory, matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies \citep{Lai96}. This assumption is generally valid if the length scales of interest are large compared with the length scales of discrete molecular structure, but whether the approximation of continuum mechanics is actually justified in a given situation is merely a matter of experimental test.

Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different sollicitations that occur during medical procedures. A sufficient knowledge of continuum mechanics is therefore essential to follow the rest of this manuscript. Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials. This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models. This chapter will follow the notation used by \cite{Bonnet97} and \cite{Reddy07}. The interested reader may refer to these books for more details.

\section{Description of motion}
Let us consider a body $\boldsymbol \beta$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\boldsymbol \beta$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. Deformation implies the change in shape and\slash or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.

Let us now consider a given particule of the body that we call $X$. What we will call particule in the following is in fact an infinitesimal volume of material. We denote the position it occupies in the initial configuration $\mathbf{X}$ and note $\mathbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference. The mapping $\chi$ defined as the following:
\begin{equation}
\label{chap2:mapping}
\chi : \boldsymbol \beta_{\kappa_{0}} \rightarrow \boldsymbol \beta_{\kappa}
\end{equation} 
is called the \emph{deforming mapping} of the body $\boldsymbol \beta$ from $\kappa_{0}$ to $\kappa$. When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time. Its mathematical description follows one of the two approaches: the material description or the spatial description. The material description is known as \emph{Lagrangian description} whereas the spatial description is also called \emph{Eulerian description}. These two approaches are detailled next.

@@

	\subsection{Eulerian description}
Rather than following the particules, the Eulerian description observes the changes at fixed locations. 	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with different material particules at different times, because different particules occupy the position $\textbf{x} \in \kappa$ at different times. A change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule. Hence, the Eulerian description is focused on a spatial position. 	This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

	\subsection{Displacement field}
The deplacement of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is the following:
\begin{equation}
\textbf{u}(\textbf{X}, t) = \textbf{x}(\textbf{X}, t) - \textbf{X}.
\end{equation}
A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$. Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.

@@

Consider two particules $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
\begin{equation}
\label{chap2:dX}
d\mathbf{X} = \mathbf{X}_{P_{2}} - \mathbf{X}_{P_{1}}.
\end{equation}
After deformation the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
\begin{equation}
\mathbf{x}_{P_1} = \chi(\mathbf{X}_{P_{1}}, t) \quad \mbox{and} \quad \mathbf{x}_{P_2} = \chi(\mathbf{X}_{P_{2}}, t).
\end{equation}
Using \eqref{chap2:dX} the distance $d\textbf{x}$ between $P_{1}$ and $P_{2}$ can then be expressed as:
\begin{equation}
d\mathbf{x} = \mathbf{x}_{P_{2}} - \mathbf{x}_{P_{1}} = \chi(\mathbf{X}_{P_{1}} + d\mathbf{X}, t) - \chi(\mathbf{X}_{P_{1}}, t).
\end{equation}
The \emph{deformation gradient} $\textbf{F}$ can be defined as:
\begin{equation}
\mathbf{F} = \pd{\chi}{\mathbf{X}}
\end{equation}
and the vector $d\textbf{x}$ can then be obtained in terms of $d\mathbf{X}$ as:
\begin{equation}
\label{chap2:dxdX}
d\textbf{x} = \textbf{F} \, d\textbf{X}.
\end{equation}
Note that $\textbf{F}$ transforms vectors from the reference configuration into vectors in the current configuration and is therefore a second-order tensor.

@@

\section{Strain measures}
The length of a material curve from the reference configuration can change when displaced to a curve in the deformed configuration. If all the curves do not change length, it is said that a rigid body displacement occurred. The concept of strain is used to evaluate how much a given displacement differs locally from a rigid body displacement. Therefore, although we know how to transform vectors from the reference configuration into vectors in the current configuration using the deformation gradient, it is more useful to find a measure of the change in length of $d\mathbf{X}$. Many measures of strains can be defined and the most common ones are now introduced.

	\subsection{Cauchy-Green deformation tensors}
Consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration. In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$. We are interested in the change of distance between the two points $P_{1}$ and $P_{2}$ as the body deforms.

@@

	\subsection{Green strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-St. Venant strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green strain tensor is a symmetric second-order tensor. Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be developped as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right] \label{chap2:GreenTensor}
\end{align}

@@

	\subsection{Principal strains and invariants}	\label{chap2:strainInvariants}
The components $\varepsilon_{ij}$ of the strain tensor depend on of the coordinate system at the point under consideration. However, the strain tensor itself is a physical quantity and as such, it is independent of the coordinate system chosen to represent it. Certain operations on strain tensors give the same result independently of the coordinate system chosen to represent the components of strain. As an example, a vector is a simple tensor of rank one. In three dimensions, it has three components. The value of these components will depend on the coordinate system chosen to represent the vector, but the length of the vector is a physical quantity (a scalar) and is independent of the coordinate system chosen to represent the vector. Similarly, every second rank tensor (such as the strain tensors) has three independent invariant quantities associated with it. One set of such \emph{invariants} are the principal strains of the strain tensor, which are just the eigenvalues of the strain tensor. Because they are independent of any coordinate system, they are very convenient to define strain energy density functions (see section \ref{chap2:non-linear}). The most commonly used invariants are:
\begin{align}
I &= tr(\varepsilon) \notag \\
II &= \frac{1}{2} \left\lbrace tr(\varepsilon^{2}) - \left[tr(\varepsilon)^{2} \right] \right\rbrace \\
III &= det(\varepsilon). \notag
\end{align}

@@

		\subsection{Principal stresses and invariants}
In the same way that we defined principal strains in section \ref{chap2:strainInvariants} and like any second-rank tensor, the stress tensor has three invariants and we can find a coordinate system $(\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3)$ in which the expression of the stress tensor is:
\begin{equation}
\boldsymbol \sigma =
	\begin{bmatrix}
	\sigma_1 &         0      &        0       \\
	        0      & \sigma_2 &        0       \\
	        0      &          0     & \sigma_3
	\end{bmatrix}	
\end{equation}
This result means that at every point in a stressed body there are at least three planes called \emph{principal planes}, with normal vectors $\mathbf{n}_1$, $\mathbf{n}_2$ and $\mathbf{n}_3$ called \emph{principal directions} (in fact, the eigenvectors), where the corresponding stress vector is perpendicular to the plane and where there are no normal shear stresses. The three stresses normal $\sigma_1$, $\sigma_2$ and $\sigma_3$ to these principal planes are called \emph{principal stresses}.

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, no mention was made of any material in the derivations. Therefore these relations are not sufficient to describe the response of a material to a given loading. We know that the stresses result from the deformation of the material and it is now necessary to express them in terms of some measure of this deformation such as, for instance, the strain. These relationships, known as \emph{constitutive equations}, obviously depend on the type of material under consideration and may be dependent on time or not. Constitutive equations are not derived from any physical principles, although they are subject to obeying certain rules and the entropy inequality. In essence, constitutive equations are mathematical models of the behaviour of materials that are validated against experimental results.

@@

	\subsection{Elasticity}	
If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material elastic. The work done by the stress is generally dependent on the deformation path. However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation. It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.

@@

Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young's modulus, shear modulus, and so on. These constants are measured using simple tests like uniaxial tension test or pure shear test. Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$. Therefore, we will now reformulate \eqref{chap2:orthotropic} with respect to those engineering constants. Nothing $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:

@@

\begin{equation}
\label{chap2:orthotropic}
	\begin{bmatrix}
		\varepsilon_{11} \\
		\varepsilon_{22} \\
		\varepsilon_{33} \\
		\varepsilon_{23} \\
		\varepsilon_{13} \\
		\varepsilon_{12}
	\end{bmatrix}
= 
	\begin{bmatrix}
		\frac{1}{E_1} & -\frac{\nu_{21}}{E_2} & -\frac{\nu_{31}}{E_3} &       0       &       0       &       0 \\
		-\frac{\nu_{12}}{E_1} & \frac{1}{E_2} & -\frac{\nu_{32}}{E_3} &       0       &       0       &       0 \\
		-\frac{\nu_{13}}{E_1} & -\frac{\nu_{23}}{E_2} & \frac{1}{E_3} &       0       &       0       &       0 \\
		      0       &        0     &        0       & \frac{1}{G_{23}} &       0       &       0 \\
		      0       &        0     &        0       &       0       & \frac{1}{G_{13}}&       0 \\
		      0       &        0     &        0       &       0       &       0       & \frac{1}{G_{12}}
	\end{bmatrix}
	\begin{bmatrix}
		\sigma_{11} \\
		\sigma_{22} \\
		\sigma_{33} \\
		\sigma_{23} \\
		\sigma_{13} \\
		\sigma_{12}
	\end{bmatrix}
	.
\end{equation}

	\subsection{Isotropic materials} \label{chap2:isotropic}
A material is said \emph{isotropic} if its properties are independent of the directions. Therefore, we have the following relations:
\begin{equation}
E_1 = E_2 = E_3 = E, \quad G_{12} = G_{13} = G_{23} = G, \quad \nu_{12} = \nu_{23} = \nu_{13} = \nu.
\end{equation}
The stress-strain relations may be written:
\begin{equation}
\sigma_{ij} = \frac{E}{1+\nu} \varepsilon_{ij} + \frac{\nu E}{(1+\nu)(1-2\nu)} \varepsilon_{kk}\delta_{ij}.
\end{equation}	
where the summation of repeated indices is implied (Einstein's notation). The inverse relations are:
\begin{equation}
\varepsilon_{ij} = \frac{1+\nu}{E} \sigma_{ij} - \frac{\nu}{E} \sigma_{kk} \delta_{ij},
\end{equation}
recalling that $G = E/[2(1+\nu)]$.

@@

	\subsubsection*{Mooney-Rivlin}
The Mooney-Rivlin model was proposed by Melvin Mooney and Ronald Rivlin in two independent papers in 1952. For an incompressible Mooney-Rivlin material, the strain energy density function $U_0$ is taken as a linear function of the principal invariants of the left Cauchy-Green strain tensor $\mathbf{B}$ and is written as:
\begin{equation}
U_0 = C_1 (I_B-3) + C_2(II_B-3),
\end{equation}
where $C_1$ and $C_2$ are two constants and $I_B$ and $II_B$ are the two principal invariants of $\mathbf{B}$. For those materials, it can be shown that the stress tensor has the form
\begin{equation}
\sigma = -p \mathbf{I} + \alpha \mathbf{B} + \beta \mathbf{B}^{-1},
\end{equation}
where $\alpha$ and $\beta$ are given by
\begin{equation}
\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2 \quad \mbox{with } \rho \, \mbox{the material density.}
\end{equation}
The Mooney-Rivlin incompressible model is often used to represent the behaviour of rubber-like materials.

>>>>

\chapter{Background in continuum mechanics for soft-tissue modelling}
\label{chap:softtissue}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

\section{Introduction}
In our everyday life, matter appears smooth and continuous: from the wood used to build your desk to the water you drink. But this is just illusion. The concept that matter is composed of discrete units has been around for millennia. In fact, we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century \citep{Lautrup05}. However, certain physical phenomena can be predicted with theories that pay no attention to the molecular structure of materials. Consider for instance the deformation of the horizontal board of a bookshelf under the weight of books. The bending of the shelf can be modelled without considering its molecular composition. The branch of physics in which materials are treated as continuous is known as \emph{continuum mechanics}. Continuum mechanics studies the response of materials to different loading conditions. In this theory, matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies \citep{Lai96}. This assumption is generally valid if the length scales of interest are large compared with the length scales of discrete molecular structure, but whether the approximation of continuum mechanics is actually justified in a given situation is merely a matter of experimental test.

Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different sollicitations that occur during medical procedures. A sufficient knowledge of continuum mechanics is therefore essential to follow the rest of this manuscript. Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials. This chapter will not only deal with those two aspects but will also discuss the different aspects of the mechanical behaviour that we find in anatomical structures. Note that this chapter will mostly follow the notation used by \cite{Bonnet97} and \cite{Reddy07}. The interested reader may refer to these books for more details.  
%This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.

\section{Description of motion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. Deformation implies the change in shape and\slash or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.

Let us now consider a given particule of the body that we call $X$. What we will call particule in the following is in fact an infinitesimal volume of material. We denote the position it occupies in the initial configuration $\mathbf{X}$ and note $\mathbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference. The mapping $\chi$ defined as the following:
\begin{equation}
\label{chap2:mapping}
\chi : \textbf{\textit{B}}_{\kappa_{0}} \rightarrow \textbf{\textit{B}}_{\kappa}
\end{equation} 
is called the \emph{deforming mapping} of the body $\textbf{\textit{B}}$ from $\kappa_{0}$ to $\kappa$. When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time. Its mathematical description follows one of the two approaches: the material description or the spatial description. The material description is known as \emph{Lagrangian description} whereas the spatial description is also called \emph{Eulerian description}. These two approaches are detailled next.

@@

	\subsection{Eulerian description}
Rather than following the particules, the Eulerian description observes the changes at fixed locations. 	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particule occupying the position $\textbf{x} \in \kappa$, which may very well be a different particule for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

	\subsection{Displacement field}
The displacement $ \mathbf{u} $ of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
\begin{equation}
\mathbf{u}(\textbf{X}, t) = \mathbf{x}(\mathbf{X}, t) - \mathbf{X}.
\end{equation}
A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$. Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.

@@

Consider two particules $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
\begin{equation}
\label{chap2:dX}
d\mathbf{X} = \mathbf{X}_{P_{2}} - \mathbf{X}_{P_{1}}.
\end{equation}
After deformation, the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
\begin{equation}
\mathbf{x}_{P_1} = \chi(\mathbf{X}_{P_{1}}, t) \quad \mbox{and} \quad \mathbf{x}_{P_2} = \chi(\mathbf{X}_{P_{2}}, t).
\end{equation}
Using \eqref{chap2:dX} the distance $d\textbf{x}$ between $P_{1}$ and $P_{2}$ can then be expressed as:
\begin{equation}
d\mathbf{x} = \mathbf{x}_{P_{2}} - \mathbf{x}_{P_{1}} = \chi(\mathbf{X}_{P_{1}} + d\mathbf{X}, t) - \chi(\mathbf{X}_{P_{1}}, t).
\end{equation}
The \emph{deformation gradient} $\textbf{F}$ can be defined as:
\begin{equation}
\mathbf{F} = \pd{\chi}{\mathbf{X}}
\end{equation}
and the vector $d\textbf{x}$ can then be obtained in terms of $d\mathbf{X}$ as:
\begin{equation}
\label{chap2:dxdX}
d\textbf{x} = \textbf{F} \, d\textbf{X}.
\end{equation}
Note that $\textbf{F}$ transforms vectors from the reference configuration into vectors in the current configuration and is therefore a second-order tensor.

@@

\section{Strain measures}
The length of a material curve from the reference configuration can change when displaced to a curve in the deformed configuration. If all the curves do not change length, it is said that a rigid body displacement occurred. The concept of strain is used to evaluate how much a given displacement differs locally from a rigid body displacement. Therefore, although we know how to transform vectors from the reference configuration into vectors in the current configuration using the deformation gradient, it is more useful to find a measure of the change in length of $d\mathbf{X}$. Many measures of strains can be defined and the most common ones will now be introduced.

	\subsection{Cauchy-Green deformation tensors}
Let us consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration. In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$. We are interested in the change of distance between the two points $P_{1}$ and $P_{2}$ as the body deforms.

@@

	\subsection{Green strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-St.\,Venant strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green strain tensor is a symmetric second-order tensor. Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be developped as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right] \label{chap2:GreenTensor}
\end{align}

@@

	\subsection{Principal strains and invariants}	\label{chap2:strainInvariants}
The components $\varepsilon_{ij}$ of the strain tensor depend on the coordinate system at the point under consideration. However, the strain tensor itself is a physical quantity and as such, it is independent of the coordinate system chosen to represent it. Certain operations on strain tensors give the same result independently of the coordinate system chosen to represent the components of strain. As an example, a vector is a simple tensor of rank one. In three dimensions, it has three components. The value of these components will depend on the coordinate system chosen to represent the vector, but the length of the vector is a physical quantity (a scalar) and is independent of the coordinate system chosen to represent the vector. Similarly, every second rank tensor (such as the strain tensors) has three independent invariant quantities associated with it. One set of such \emph{invariants} are the principal strains of the strain tensor, which are just the eigenvalues of the strain tensor. Because they are independent of any coordinate system, they are very convenient to define strain energy density functions (see section \ref{chap2:non-linear}). The most commonly used invariants are:
\begin{align}
I_1 &= tr(\varepsilon) \notag \\
I_2 &= \frac{1}{2} \left\lbrace tr(\varepsilon^{2}) - \left[tr(\varepsilon)\right]^{2} \right\rbrace \\
I_3 &= det(\varepsilon). \notag
\end{align}

@@

		\subsection{Principal stresses and invariants}
In the same way that we defined principal strains in section \ref{chap2:strainInvariants} and like any second-rank tensor, the stress tensor has at least three invariants and we can find a coordinate system $(\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3)$ in which the expression of the stress tensor is:
\begin{equation}
\boldsymbol \sigma =
	\begin{bmatrix}
	\sigma_1 &         0      &        0       \\
	        0      & \sigma_2 &        0       \\
	        0      &          0     & \sigma_3
	\end{bmatrix}	
\end{equation}
This result means that at every point in a stressed body there are at least three planes called \emph{principal planes}, with normal vectors $\mathbf{n}_1$, $\mathbf{n}_2$ and $\mathbf{n}_3$ called \emph{principal directions} (in fact, the eigenvectors), where the corresponding stress vector is perpendicular to the plane and where there are no normal shear stresses. The three stresses normal $\sigma_1$, $\sigma_2$ and $\sigma_3$ to these principal planes are called \emph{principal stresses}.

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, no mention was made of any material in the derivations. Therefore, these relations are not sufficient to describe the response of a material to a given loading. We know that the stresses result from the deformation of the material and it is now necessary to express them in terms of some measure of this deformation such as, for instance, the strain. These relationships, known as \emph{constitutive equations}, obviously depend on the type of material under consideration and may be dependent on time or not. Constitutive equations are not derived from any physical principles, although they are subject to obeying certain rules and the entropy inequality. In essence, constitutive equations are mathematical models of the behaviour of materials that are validated against experimental results.

@@

	\subsection{Elasticity}	
If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}. The work done by the stress is generally dependent on the deformation path. However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation. It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.

@@

Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young's modulus, shear modulus, and so on. These constants are measured using simple tests like uniaxial tension test or pure shear test. Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$. Therefore, we will now reformulate \eqref{chap2:orthotropic} with respect to those engineering constants. Noting $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:

@@

\begin{equation}
\label{chap2:orthotropic2}
	\begin{bmatrix}
		\varepsilon_{11} \\
		\varepsilon_{22} \\
		\varepsilon_{33} \\
		\varepsilon_{23} \\
		\varepsilon_{13} \\
		\varepsilon_{12}
	\end{bmatrix}
= 
	\begin{bmatrix}
		\dfrac{1}{E_1} & -\dfrac{\nu_{21}}{E_2} & -\dfrac{\nu_{31}}{E_3} &       0       &       0       &       0 \\
		-\dfrac{\nu_{12}}{E_1} & \dfrac{1}{E_2} & -\dfrac{\nu_{32}}{E_3} &       0       &       0       &       0 \\
		-\dfrac{\nu_{13}}{E_1} & -\dfrac{\nu_{23}}{E_2} & \dfrac{1}{E_3} &       0       &       0       &       0 \\
		      0       &        0     &        0       & \dfrac{1}{G_{23}} &       0       &       0 \\
		      0       &        0     &        0       &       0       & \dfrac{1}{G_{13}}&       0 \\
		      0       &        0     &        0       &       0       &       0       & \dfrac{1}{G_{12}}
	\end{bmatrix}
	\begin{bmatrix}
		\sigma_{11} \\
		\sigma_{22} \\
		\sigma_{33} \\
		\sigma_{23} \\
		\sigma_{13} \\
		\sigma_{12}
	\end{bmatrix}
	.
\end{equation}

	\subsection{Isotropic materials} \label{chap2:isotropic}
A material is said \emph{isotropic} if its properties are independent of directions. Therefore, we have the following relations:
\begin{equation}
E_1 = E_2 = E_3 = E, \quad G_{12} = G_{13} = G_{23} = G, \quad \nu_{12} = \nu_{23} = \nu_{13} = \nu.
\end{equation}
The stress-strain relations may be written:
\begin{equation}
\sigma_{ij} = \frac{E}{1+\nu} \varepsilon_{ij} + \frac{\nu E}{(1+\nu)(1-2\nu)} \varepsilon_{kk}\delta_{ij}.
\end{equation}	
where the summation of repeated indices is implied (Einstein's notation). The inverse relations are:
\begin{equation}
\varepsilon_{ij} = \frac{1+\nu}{E} \sigma_{ij} - \frac{\nu}{E} \sigma_{kk} \delta_{ij},
\end{equation}
recalling that $G = E/[2(1+\nu)]$.

@@

	\subsubsection*{Mooney-Rivlin}
The Mooney-Rivlin model was proposed by Melvin Mooney and Ronald Rivlin in two independent papers in 1952. For an incompressible Mooney-Rivlin material, the strain energy density function $U_0$ is taken as a linear function of the principal invariants of the left Cauchy-Green strain tensor $\mathbf{B}$ and is written as:
\begin{equation}
U_0 = C_1 (I_B-3) + C_2(II_B-3),
\end{equation}
where $C_1$ and $C_2$ are two constants and $I_B$ and $II_B$ are the two principal invariants of $\mathbf{B}$. For those materials, it can be shown that the stress tensor has the form
\begin{equation}
\sigma = -p \mathbf{I} + \alpha \mathbf{B} + \beta \mathbf{B}^{-1},
\end{equation}
where $\rho$ is the material density and $\alpha$ and $\beta$ are given by
\begin{equation}
\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2
\end{equation}
The Mooney-Rivlin incompressible model is often used to represent the behaviour of rubber-like materials.
